---
title: "p8105_hw2_dfk2117"
output: "github_document"
date: "2023-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required libraries
```{r, echo=FALSE}
library(tidyverse)
```


## Problem 1

### Wotrking with the politicians data 

```{r}
df_politicians =
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day")) |> 
  mutate(month = month.name[as.numeric(month)]) |> 
  mutate(
    president = case_when(
      prez_gop %in% c(1,2) ~ "gop",
      prez_dem == 1 ~ "dem",
    )
  ) |>  
  select(-prez_gop, -prez_dem, -day) |> 
  arrange(year, month)
```

### Working with the S&P data

```{r}
df_snp = 
  read_csv("data/snp.csv") |> 
  janitor::clean_names() |> 
  separate(date, into = c("year", "month", "day")) |> 
  mutate(month = month.name[as.numeric(month)]) |> 
  select(-day) |> 
  arrange(year, month)
```

### Unemployment data

```{r}
df_unemployment =
  read.csv("data/unemployment.csv") |> 
  janitor::clean_names()
```

### merging datasets


# Problem 2 


```{r}
df_mr_trash =
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 1) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  select(-x15, -x16) |> 
  mutate(homes_powered = ((weight_tons * 500)/30)) |> 
  mutate(trash_wheel = "Mr_Trash") |> 
  mutate(year = as.numeric(year))


df_pr_trash =
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 2) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(homes_powered = ((weight_tons * 500)/30)) |> 
  mutate(trash_wheel = "Prof_Trash") |> 
  mutate(year = as.numeric(year))

df_gw_trash =
  readxl::read_excel("data/202207 Trash Wheel Collection Data.xlsx", sheet = 4) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(homes_powered = ((weight_tons * 500)/30)) |> 
  mutate(trash_wheel = "Gwynnda") |> 
  mutate(year = as.numeric(year))

df_all_wheels = 
  bind_rows(df_gw_trash, df_mr_trash, df_pr_trash) |> 
  relocate(homes_powered) |> 
  relocate(dumpster) |> 
  relocate(trash_wheel)
  

```

Each of these three datasets contained a month, year and date variable (yyyy/mm/dd). They also contained a dumpster number variable and another two variables representing the amount of trash (in tons) and the volume of the trash (cubic yards) respectively. Each trash wheel also had counts of how many of a certain type of trash was picked up on that date. For Mr. Trash Wheel, this included Plastic Bottles, Polystyrene, Cigarette Butts, Glass Bottles	Grocery Bags, Chip Bags and Sports Balls. Professor Trash Wheel's data included all of these same counts except for the Sports Balls count. Gwynda only included 4 of the 7 counts: Plastic Bottles, Polystyrene, Cigarette Butts and Plastic Bags. Finally, all 3 datasets contained a homes_powered variable which was incomplete.


The resulting dataset contains all of the variables that the previous three contained, but now has a variable that labels which trash wheel the data comes from as well as the edited `homes_powered` variable that I made. I considered removing the counts of trash items that did not overlap in all 3 datasets, but I decided to leave them in for completeness.

## Problem 3

```{r}
df_baseline_mci =
  read_csv("data/data_mci/MCI_baseline_trimmed.csv") |>
  janitor::clean_names() |> 
  mutate(apoe4 =
    case_when(
      apoe4 == 1 ~ "carrier",
      apoe4 == 0 ~ "non-carrier")) |> 
  mutate(sex =
    case_when(
      sex == 1 ~ "male",
      sex == 0 ~ "female")) |> 
  filter(age_at_onset != ".") |> 
  mutate(age_at_onset = as.numeric(as.character(age_at_onset)))


```

During the import process, the first row needed to be removed as the names of each individual variable were being treated as an individual. I made a copy and removed the extraneous row before importing. 

Based on the dataset, 483 were recruited and of those, 97 developed mci. The mean baseline age is 70.3 years and the code that gives it is:

```{r}
df_baseline_mci |> 
  summarise(mean(age_at_onset))
```

0.652 or 65.2% of females in the baseline study are carriers of apoe4. The code for this result is below:

```{r}
df_baseline_mci |> 
  summarise(
    total_females = sum(sex == "female"),
    female_carriers = sum(sex == "female" & apoe4 == "carrier")) |> 
  mutate(f_carrier_ratio = female_carriers / total_females) |> 
  pull(f_carrier_ratio)
```




Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.





